### 2025-02-28 sports-cognitive-psychology.qmd
20250227 - edits
20250227 - clean up
20250227 - update links

Much of the history of analytics in sports traces back to Bill James.
James was a member of the Society for American Baseball Research (SABR).
He published a series of Baseball Abstracts, which included players' statistics.
James published the first Baseball Abstract in 1977 (<https://www.pbs.org/thinktank/transcript1197.html>; archived at <https://www.pbs.org/thinktank/transcript1197.html>).
He coined the term Sabermetrics (which was originally named SABRmetrics, based on the name of the society) to refer to advanced metrics and statistical/empirical analysis of baseball.
James believed that traditional statistics like batting average, runs batted in (RBIs), and pitchers' wins did not reflect a player's true value—i.e., their contribution to the team's success.
Instead, he advocated for the use of more advanced metrics that could provide a deeper understanding of a player's performance and value such as on-base percentage (OBP), slugging percentage (SLG), wins above replacement (WAR), and fielding independent pitching (FIP).
He also developed advanced metrics such as runs created (RC) and defense efficiency rating (DER).

@@

It had been known for a long time that going for it on fourth down would frequently increase a team's chances of winning.
Despite that, historically, teams rarely went for it on fourth down and elected to punt or kick a field goal instead.
It is curious that there was such a discrepancy between the decisions that would maximize teams' winning percentage and the decisions coaches actually made.
One potential explanation for the discrepancy is because of coaches' risk aversion.
As noted in @sec-cognitiveBiasesRiskAversion, when it is possible to experience either a gain or a loss from a decision, [loss aversion bias](#sec-cognitiveBiasesLossAversion) tends to lead people to make risk-averse decisions [@Kahneman2011].
According to this idea, in the case of failing to successfully convert on fourth down, coaches do not want to have to defend their decision to go for it to the media or the owner or general manager.

@@

Adapted from Ben Baldwin: https://www.nfl4th.com/articles/4th-down-research.html (archived at <https://perma.cc/S5D8-3NCU>)

>>>>

Much of the history of analytics in sports traces back to Bill James.
James was a member of the Society for American Baseball Research (SABR).
He published a series of Baseball Abstracts, which included players' statistics.
James published the first Baseball Abstract in 1977 (<https://www.pbs.org/thinktank/transcript1197.html>; archived at <https://perma.cc/Y7J7-GB9V>).
He coined the term Sabermetrics (which was originally named SABRmetrics, based on the name of the society) to refer to advanced metrics and statistical/empirical analysis of baseball.
James believed that traditional statistics like batting average, runs batted in (RBIs), and pitchers' wins did not reflect a player's true value—i.e., their contribution to the team's success.
Instead, he advocated for the use of more advanced metrics that could provide a deeper understanding of a player's performance and value such as on-base percentage (OBP), slugging percentage (SLG), wins above replacement (WAR), and fielding independent pitching (FIP).
He also developed advanced metrics such as runs created (RC) and defense efficiency rating (DER).

@@

It had been known for a long time that going for it on fourth down would frequently increase a team's chances of winning.
Despite that, historically, teams rarely went for it on fourth down and elected to punt or kick a field goal instead.
It is curious that there was such a discrepancy between the decisions that would maximize teams' winning percentage and the decisions coaches actually made.
One potential explanation for the discrepancy is because of coaches' risk aversion.
As noted in @sec-cognitiveBiasesRiskAversion, when it is possible to experience either a gain or a loss from a decision, [loss aversion bias](#sec-cognitiveBiasesLossAversion) tends to lead people to make risk-averse decisions [@Kahneman2011].
According to this idea, in the case of failing to successfully convert on fourth down, coaches do not want to have to defend their decision to go for it to the media or the owner or general manager.
That is, they may often play *not to lose*, rather than to win, in order to keep their job.

@@

Adapted from Ben Baldwin: <https://www.nfl4th.com/articles/4th-down-research.html> (archived at <https://perma.cc/S5D8-3NCU>)
